---
title: "Tools installation for seasonnal adjustment"
author: "Tanguy BARTHELEMY"
date: "2023-01-17"
output:
  github_document: default
  html_document:
    df_print: paged
    number_sections: TRUE
    highlight: "arrow"
    theme: "readable"
  pdf_document: 
    includes:
        in_header: 
            - ../xelatex-emoji/xelatexemoji.sty
    extra_dependencies: ["url"]
    number_sections: TRUE
    highlight: "arrow"
    latex_engine: xelatex
always_allow_html: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{css, echo = FALSE}
.Code {
background-color: #F1F3F5;
}
```

```{r, echo = FALSE, eval = TRUE}
html_color <- "#000cff"
windows_path_color <- "#006817"

colFmt <- function(x, color, url = TRUE) {
    color_name <- deparse(substitute(color))
    outputFormat <- knitr::opts_knit$get("rmarkdown.pandoc.to")
    x <- x |> 
        gsub(pattern = "\\", replacement = "\\\\", fixed = TRUE)
    
    if (outputFormat == 'latex') {
        x <- x |> 
            gsub(pattern = "#", replacement = "\\#", fixed = TRUE)
        ret <- paste("\\textcolor{", color_name, ifelse(url, "}{\\url{", "}{\\nolinkurl{"), x, ifelse(url, "}}", "}}"), sep = "")
    } else if (outputFormat == 'html') {
        ret <- paste("<font color='", color, "'>", x, "</font>", sep = "")
    } else {
        ret <- x
    }
    
    return(ret)
}
```

```{r, echo = FALSE, eval = TRUE}
html_color <- "#000cff"
windows_path_color <- "#006817"
```

\definecolor{html_color}{HTML}{000cff}
\definecolor{windows_path_color}{HTML}{006817}

\renewcommand{\xelatexemojipath}[1]{../xelatex-emoji/images/#1}


# Context {-}

For seasonal adjustment we recommend using JDemetra+ algorithms

Thus, it is helpful to install the following tools: 

-   JDemetra+ Graphical User Interface (X13-Arima and Tramo-Seats )

⚠️ No assistance in **SAS** language will be given. ⚠️


# Preliminary {-}

On the computers without administrator rights (*professional computer, for example*), it is recommended to create a folder **Software** under `r colFmt("C:\\Users\\...\\Software", windows_path_color, url = FALSE)` or directly under `r colFmt("C:\\Users\\Software", windows_path_color, url = FALSE)` where all software will be installed.

⚠️ Warning: when we specify an **absolute** path for a software (JDemetra+, Java, ...) in a program, a shortcut, a variable, ..., it must be modified each time any root repository is moved.

# Installation of [JDemetra+](https://github.com/jdemetra/jdemetra-app)

JDemetra+ is a collection of Java programs used for time series analysis and more specifically for seasonal adjustment. JDemetra+ is delivered in the form of an GUI (Graphical User Interface) but there is also a cruncher (executable).

## Version of JDemetra+ and dependencies

JDemetra+ is downloadable from the [github link](https://github.com/jdemetra/jdemetra-app/releases) of the application: `r colFmt("https://github.com/jdemetra/jdemetra-app/releases", html_color)`.

The last release ([v2.2.4](https://github.com/jdemetra/jdemetra-app/releases/tag/v2.2.3)) dates from january 31, 2023. It is the last **stable** version of JDemetra+. **This version should be downloaded and must be used in production.**

There is one more version of JDemetra+ which are only at a **test** stage:

-   [v3.0.0](https://github.com/jdemetra/jdemetra-app/releases/tag/v3.0.0-RC1): the new JDemetra+ version with new features and a new GUI

JDemetra+ v2.2.4 require Java version $\geq 8$ while v3.0.0 requires Java version $\geq 17$:

| JDemetra+ version | Java version |
|-------------------|--------------|
|            v2.2.4 |        $\geq 8$ |
|      v3.0.0 (RC1) |       $\geq 17$ |

*RC: Release candidate*

In the following procedure, the installation processes of this 3 versions are the same. You just have to repeat them for each version you want to install.


## Installation process

There are two possibilities for installing:

-   **Download and execute** the `.exe` file which requires administrator rights
-   **Download** and **unzip** the compressed folder `.zip` that allows to get a portable version of the software

⚠️ Warning: for the second option, you need to **download** the compressed folder `jdemetra+-2.2.4-bin.zip` (for the version 2.2.4 for example) and **not** the folder `Source code (zip)`.

The Software is in the folder `r colFmt("\\nbdemetra\\bin\\", windows_path_color, url = FALSE)`, these are the file `nbdemetra.exe` (version 32-bit) and `nbdemetra64.exe` (version 64-bit).

ℹ Advice: If you want to use several versions of JDemetra+ (v2.2.4, v3.0.0, ...), you can rename the unzipped folder in `r colFmt("\\nbdemetra-2.2.4\\", windows_path_color, url = FALSE)` and `r colFmt("\\nbdemetra-3.0.0\\", windows_path_color, url = FALSE)`.

ℹ️ Remark:You can create shortcuts to the executable files if you want to launch them from another folder (Desktop, project folder...).

## Installation of the cruncher

The cruncher ([**JWSACruncher**](https://github.com/jdemetra/jwsacruncher)) is a tool to update a workspace of JDemetra+ from the console, without opening JDemetra+ Graphical User Interface. The update of a workspace can then be done from another Software (**R** or **SAS** for example).

To use the cruncher, you have to:

-   **Download** and **unzip** the file from the **latest stable** version v2.2.4 here `r colFmt("https://github.com/jdemetra/jwsacruncher/releases", html_color)`

If you want to install and use a portable Java version (See section [Java installation](#install_java)), you have to modify some parameters to use the cruncher:

-   In the unzipped folder, **open** (for example with Notepad++) the file `jwsacruncher.bat` present in the subfolder `r colFmt("\\bin\\", windows_path_color, url = FALSE)` (that is under `r colFmt("jdemetra-cli-2.2.4\\bin\\", windows_path_color, url = FALSE)` in the version 2.2.3 of the cruncher)
-   **Modify** the value of the variable `JAVACMD` at the line **71** (currently `JAVACMD=java`) by the address towards the file `java.exe` of the portable version. Then, if JPortable is installed under `r colFmt("C:\\Users\\Software", windows_path_color, url = FALSE)`, the new line is `if "%JAVACMD%"=="" set JAVACMD="C:\\Users\\Software\\Java64\\bin\\java"` (for Java 8).


# Installation of Java {#install_java}

ℹ️ On Insee computers, Java is already installed in version 8. Then, there is no need to install a portable version to use JDemetra+ in version 2.2.4.

## Java 8

To install Java 8, use the link `r colFmt("https://portableapps.com/apps/utilities/java_portable", html_color)`. If you use the version 64-bit of JDemetra+, you should install the version jPortable 64-bit (at the bottom of the page).


## Java 17

To install Java 17, you need to head over to `r colFmt("https://www.oracle.com/java/technologies/javase/jdk17-archive-downloads.html", html_color)`.

-   **Download** the version `Compressed Archive` of Windows (`r colFmt("https://download.oracle.com/java/17/archive/jdk-17.0.4.1_windows-x64_bin.zip", html_color)`)
-   **Unzip** the folder `r colFmt("jdk-17.0.6", windows_path_color, url = FALSE)` under `r colFmt("C:\\Users\\Software", windows_path_color, url = FALSE)` (for example)

After a Java installation (in version 8, 17 or other), you need to:

- **Modify** the environment variable `PATH` of Windows
- **Modify** the targets of JDemetra+ to set the localisation of the new Java versions.

For example, if you installed Java version 17 in order to use the JDemetra+ version 3.0.0. You should add the localisation of Java 17 to the shortcut of the executable using the option `--jdkhome`. The target of the shortcut becomes `C:\Users\Software\nbdemetra-3.0.0\bin\nbdemetra64.exe --jdkhome "C:\Users\Software\Java17\jdk17"`
