---
title: "Tools installation for seasonnal adjustment"
author: "Tanguy BARTHELEMY"
date: "2023-01-17"
output:
  pdf_document: 
    includes:
        in_header: 
            - ../xelatex-emoji/xelatexemoji.sty
    extra_dependencies: ["url"]
    number_sections: TRUE
    highlight: "arrow"
    latex_engine: xelatex
  html_document:
    df_print: paged
    number_sections: TRUE
    highlight: "arrow"
    theme: "readable"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{css, echo = FALSE}
.Code {
background-color: #F1F3F5;
}
```

```{r, echo = FALSE, eval = TRUE}
html_color <- "#000cff"
windows_path_color <- "#006817"

colFmt <- function(x, color, url = TRUE) {
    color_name <- deparse(substitute(color))
    outputFormat <- knitr::opts_knit$get("rmarkdown.pandoc.to")
    x <- x |> 
        gsub(pattern = "\\", replacement = "\\\\", fixed = TRUE)
    
    if (outputFormat == 'latex') {
        x <- x |> 
            gsub(pattern = "#", replacement = "\\#", fixed = TRUE)
        ret <- paste("\\textcolor{", color_name, ifelse(url, "}{\\url{", "}{\\nolinkurl{"), x, ifelse(url, "}}", "}}"), sep = "")
    } else if (outputFormat == 'html') {
        ret <- paste("<font color='", color, "'>", x, "</font>", sep = "")
    } else {
        ret <- x
    }
    
    return(ret)
}
```

```{r, echo = FALSE, eval = TRUE}
html_color <- "#000cff"
windows_path_color <- "#006817"
```

\definecolor{html_color}{HTML}{000cff}
\definecolor{windows_path_color}{HTML}{006817}

\renewcommand{\xelatexemojipath}[1]{../xelatex-emoji/images/#1}


# Context {-}

The helping tools for seasonal adjustment are:

-   **R** and Rstudio
-   JDemetra+

⚠️ No assistance in **SAS** language will be given. ⚠️


# Preliminary {-}

On the computers without administrator rights (*professionnal computer, for example*), it is recommended to create a folder **Software** under `r colFmt("C:\\Users\\...\\Software", windows_path_color, url = FALSE)` or directly under `r colFmt("C:\\Users\\Software", windows_path_color, url = FALSE)` where all software will be installed.

⚠️ Warning: when we specify an **absolute** path for a software (JDemetra+, Java, **R**, ...) in a program, a shortcut, a variable, ..., it must be modified each time any root repository is moved.

# Installation of [JDemetra+](https://github.com/jdemetra/jdemetra-app)

JDemetra+ is a collection of Java programs used for the study of time series and more specifically for seasonnal adjustment. JDemetra+ is delivered in the forme of an GUI (Graphical User Interface) but there are **R** packages developed to be used with **R** as well as a cruncher (executable).

## Version of JDemetra+ and dependencies

JDemetra+ is downloadable from the [github link](https://github.com/jdemetra/jdemetra-app/releases) of the application: `r colFmt("https://github.com/jdemetra/jdemetra-app/releases", html_color)`.

The last release ([v2.2.3](https://github.com/jdemetra/jdemetra-app/releases/tag/v2.2.3)) dates from July 7, 2020. It is the last **stable** version of JDemetra+. **This version should be downloaded and must be used in production.**

There are two more versions of JDemetra+ which are only at the **test** stage:

-   [v2.2.4](https://github.com/jdemetra/jdemetra-app/releases/tag/v2.2.4-RC2): a v2.2.3 corrected from bugs but without adding some major new features
-   [v3.0.0](https://github.com/jdemetra/jdemetra-app/releases/tag/v3.0.0-RC1): the new JDemetra+ version with new features and a new GUI

JDemetra+ v2.2.3 and v2.2.4 require Java version $\geq 8$ while v3.0.0 require Java version $\geq 17$:

| JDemetra+ version | Java version |
|-------------------|--------------|
|            v2.2.3 |        $\geq 8$ |
|      v2.2.4 (RC2) |        $\geq 8$ |
|      v3.0.0 (RC1) |       $\geq 17$ |

*RC: Release candidate*

For the sequel, the installation processes of this 3 versions are the same. You just have to repeat them for each version you want to install.


## Installation process

There are two possibility for installing:

-   **Download and exécuter** the file `.exe` which need administrator rights
-   **Download** and **unzip** the compressed folder `.zip` that allows to get a portable version of the software

⚠️ Warning: for the second option, you need to **download** the compressed folder `jdemetra+-2.2.3-bin.zip` (for the version 2.2.3 for example) and **not** the folder `Source code (zip)`.

The Software is in the folder `r colFmt("\\nbdemetra\\bin\\", windows_path_color, url = FALSE)`, these are the file `nbdemetra.exe` (version 32-bit) and `nbdemetra64.exe` (version 64-bit).

ℹ️ Advise: If you want to use several version of JDemetra+ (v2.2.3, v3.0.0, ...), you can rename the unzipped folder in `r colFmt("\\nbdemetra-2.2.3\\", windows_path_color, url = FALSE)`, `r colFmt("\\nbdemetra-2.2.4\\", windows_path_color, url = FALSE)` and `r colFmt("\\nbdemetra-3.0.0\\", windows_path_color, url = FALSE)`.

ℹ️ Remark: You can create shortcut fo the executables if you want to launch them from another folder (Desktop, project folder...).


## Installation of the cruncher

The cruncher ([**JWSACruncher**](https://github.com/jdemetra/jwsacruncher)) is a tool to update a workspace of JDemetra+ from the console, without opening JDemetra+. The update of a workspace can then be done from another Software (**R** or **SAS** for example).

To use the cruncher, you have to:

-   **Download** and **unzip** the file from the **latest stable** version v2.2.3 here `r colFmt("https://github.com/jdemetra/jwsacruncher/releases", html_color)`

If you want to install and use a portable Java version (See section [Java installation](#install_java)), you have to modified some parameters to use the cruncher:

-   In the unziped folder, **open** (for example with Notepad++) the file `jwsacruncher.bat` present in the subfolder `r colFmt("\\bin\\", windows_path_color, url = FALSE)` (that is under `r colFmt("jdemetra-cli-2.2.3\\bin\\", windows_path_color, url = FALSE)` in the version 2.2.3 of the cruncher)
-   **Modify** the valeur of the variable `JAVACMD` at the line **71** (currently `JAVACMD=java`) by the adress toward the file `java.exe` of the portable version. Then, if JPortable is installed under `r colFmt("C:\\Users\\Software", windows_path_color, url = FALSE)`, the new line is `if "%JAVACMD%"=="" set JAVACMD="C:\\Users\\Software\\Java64\\bin\\java"` (for Java 8).


# Installation of Java {#install_java}

ℹ️ On Insee computers, Java is already installed in version 8. Then, there is no need to install a portable version to use JDemetra+ in version 2.2.3 or 2.2.4.

## Java 8

To install Java 8, use the link `r colFmt("https://portableapps.com/apps/utilities/java_portable", html_color)`. If you use the version 64-bit of JDemetra+, you should install the version jPortable 64-bit (at the bottom of the page).


## Java 17

To install Java 17, you need to go at the adresse `r colFmt("https://www.oracle.com/java/technologies/javase/jdk17-archive-downloads.html", html_color)`.

-   **Download** the version `Compressed Archive` of Windows (`r colFmt("https://download.oracle.com/java/17/archive/jdk-17.0.4.1_windows-x64_bin.zip", html_color)`)
-   **Unzip** the folder `r colFmt("jdk-17.0.6", windows_path_color, url = FALSE)` under `r colFmt("C:\\Users\\Software", windows_path_color, url = FALSE)` (for example)

After a Java installation (in version 8, 17 or other), you need to:

- **Modify** the environment variable `PATH` of Rstudio and of Windows (See section [Environment variables](#var_env))
- **Modify** the targets of JDemetra+ to inform of the localisation of the new Java versions.

For example, if you installed Java version 17 to use the JDemetra+ version 3.0.0. You should add the localisation of Java 17 at the shortcut of the executable using the option `--jdkhome`. The target of the shortcut become `C:\Users\Software\nbdemetra-3.0.0\bin\nbdemetra64.exe --jdkhome "C:\Users\Software\Java17\jdk17"`


# Installation of **R** and Rstudio

The JDemetra+ features are availables on **R** via **R** packages.
To use **R**, it is better to use an IDE so Rstudio. All the executables to download are under `r colFmt("https://posit.co/download/rstudio-desktop/#download", html_color)`.


## Installation of **R**

To install **R**, you should:

-   **Download** the binary file `R-4.2.2-win.exe` under `r colFmt("https://cran.rstudio.com/bin/windows/base/", html_color)`
-   **Execute** the executable to parameter and install **R**.


## Installation of Rstudio

**Download** the last Rstudio version (under `r colFmt("https://posit.co/download/rstudio-desktop/#download", html_color)`) and the **installer**.

If the installation via the file `.exe` fails (because it requires higher rights (administrator, elevation, ...), we will get a portable version of the Software. To do this:

- **Download** and **unzip** the compressed folder `.zip` in a folder named "Rstudio" (under `r colFmt("C:\\Users\\Software", windows_path_color, url = FALSE)`) 
- **Create a shortcut** of the file `rstudio.exe` on the Desktop.


## Installation of **R** packages

To install a **R** package, there is several methods:

- Either it is on CRAN and you can install it directly with the function `install.packages()`{.r}
- Or it is on Github and you can install it directly with the function `install_github()`{.r} from the package **remotes**
- Or you have to retrieve the package at the binary format (`.zip`) and then install with the function `install.packages()`{.r} with the arguments `repos = NULL, type = "binary"`.


### In version 2

The packages in version 2 are:

| Name | Available on CRAN | Github link |
|---|--|-----|
|    RJDemetra | ✔️ |      `r colFmt("https://github.com/jdemetra/rjdemetra", html_color)` |
| rjdworkspace | ❌ | `r colFmt("https://github.com/InseeFrLab/rjdworkspace", html_color)` |
|   JDCruncheR | ❌ |      `r colFmt("https://github.com/InseeFr/JDCruncheR", html_color)` |

The packages installation code is below:

```{r, echo = TRUE, eval = FALSE}
# If remotes is not installed
# install.packages("remotes")

install.packages("RJDemetra")
remotes::install_github("InseeFrLab/rjdworkspace")
remotes::install_github("InseeFr/JDCruncheR")
```


### In version 3

Currently no package of the version 3 are on CRAN. To install them, you need to go through Github:

```{r, echo = TRUE, eval = FALSE}
# If remotes is not installed
# install.packages("remotes")

remotes::install_github("palatej/rjdemetra3")

remotes::install_github("palatej/rjd3toolkit")
remotes::install_github("palatej/rjd3modelling")
remotes::install_github("palatej/rjd3sa")
remotes::install_github("palatej/rjd3arima")
remotes::install_github("palatej/rjd3x13")
remotes::install_github("palatej/rjd3tramoseats")
remotes::install_github("palatej/rjdemetra3")
remotes::install_github("palatej/rjdfilters")
remotes::install_github("palatej/rjd3sts")
remotes::install_github("palatej/rjd3highfreq")
remotes::install_github("palatej/rjd3stl")
remotes::install_github("palatej/rjd3bench")

remotes::install_github("AQLT/ggdemetra3")
```


### AUS case

To install a package on AUS, you can't use the function `install_github()`{.r}. Then if the package is not on CRAN, it must be downloaded at the binary format (`.zip`)

For the package **JDCruncheR**, the installing code is:

```{r, echo = TRUE, eval = FALSE}
install.packages("path/.../JDCruncheR_0.2.4.tar.gz",
                 repos = NULL, type = "binary")
```


# Environment variables {#var_env}

## In Rstudio

### Proxy

ℹ️ On Insee computers, you need to **configure** the proxy and parameters of Software localisation under Rstudio. You should:

- **Launch** the code `file.edit("~/.Renviron")`{.r}
- **Add** the parameters (news lines):

    ```{.Code}
    http_proxy = http://proxy-rie.http.insee.fr:8080/
    https_proxy = http://proxy-rie.http.insee.fr:8080/
    ```

- **Save** and **close** the file


### Java version

If a new Java version has been installed, you should inform Rstudio of the Java installation localisation. For this, as to parameter the proxy:

- **Launch** the code `file.edit("~/.Renviron")`{.r} to edit the environment variables
- **Add** the parameters:

    ```{.Code}
    JAVA_HOME = "C:/Users/Software/Java17/jdk17"
    ```

- **Save** and **close** the file

### PATH

The environment variable `PATH` in **R** is used to indicate to **R** where to find the executable files.

When you install a new software (for example JDemetra+, Rtools, Java...) that Rstudio uses, you have to modify this environment variable:

- **Get** the actual value of the variable `PATH` via the **R** command `Sys.getenv("PATH")`{.r} (Rstudio returns a succession of adresses as `C:/WINDOWS/system32;C:/WINDOWS`)

- **Copy paster** this value after `PATH = ` and add the paths towards the folder `r colFmt("\\bin\\", windows_path_color, url = FALSE)` (binary) of the software newly installed, by separating them with semicolon (without space before or after).
    For the Rtools installation, the path is `r colFmt("C:\\rtools42\\mingw64\\bin", windows_path_color, url = FALSE)` (depending on where Rtools was installed). You have to add `C:\\rtools42\\mingw64\\bin` or `C:/rtools42/mingw64/bin` (In **R**, `\` is a special character, so you have to replace the `\` by `/` or by `\\`). The path become `C:/WINDOWS/system32;C:/WINDOWS;C:/rtools42/mingw64/bin`.

- **Modify** the variable with the function `Sys.setenv()`{.r}. For Rtools, the command to launch is:

    ```{r, echo = TRUE, eval = FALSE}
    Sys.setenv(PATH = "C:/WINDOWS/system32;C:/WINDOWS;C:/rtools42/mingw64/bin")
    ```

ℹ️ NB: Generally a  32 bits version and a 64 bits version are available for downloading and installing a software. You need to check your processor type of your operating system to choose the right folder to download:

+---------------+---------------+
| Version       | Keywords      |
+===============+===============+
|               | 64            |
| 64 bits       +---------------+
|               | x64           |
+---------------+---------------+
|               | 32            |
| 32 bits       +---------------+
|               | x86           |
+---------------+---------------+

More informations on the variable `PATH` via the page `r colFmt("https://java.com/fr/download/help/path.xml", html_color)`.


# Verifications

To ensure that everything works fine, you can launch some example of **R** code and check that there is no error:

```{r, echo = TRUE, eval = FALSE}
library("RJDemetra")

myseries <- ipi_c_eu[, "FR"]
x13_model <- x13(myseries) # X-13ARIMA method
ts_model <- tramoseats(myseries) # TRAMO-SEATS method

# Basic plot with the original series, the trend and the SA series
plot(x13_model, type_chart = "sa-trend")
```

To check the Java version we are using on **R**, you can try to install and use the package **rJava** and launch the command below:

```{r, echo = TRUE, eval = FALSE}
# If rJava is not installed
install.packages("rJava")
```

If the installation of **rJava** return an error, it means that Java was incorrectly installed or incorrectly configured on **R**. You need to get back to the section [Environment variables](#var_env).

This block of code tests the Java version with which **R** works:

```{r, echo = TRUE, eval = FALSE}
library("rJava")
.jinit()
.jcall("java/lang/System", "S", "getProperty", "java.runtime.version")
```

Finally, you can consult the Java version installed with which Windows works (it doesn't matter to us):

```{r, echo = TRUE, eval = FALSE}
system("java -version")
```


# Optionnal installations

Some supplementary installations are optionnal (that is they are no mandatory but bring external features):

-   [Miktek](https://miktex.org/howto/install-miktex) to produce PDF document with Latex
-   [Rtools](https://cran.r-project.org/bin/windows/Rtools/rtools42/rtools.html) to develop **R** packages and compile the code


# Problems you may encounter

## Problems installing **R** packages

If you get the following error while installing **R** packages:

```{r, echo = TRUE, eval = FALSE}
install.packages("RJDemetra")
```

```{r, error = TRUE, echo = FALSE, include = TRUE, eval = TRUE}
stop("Erreur: the chargement a échoué
Exécution arrêtée
*** arch - x64")
```

The problem doesn't come from Java but from the **R** package. By default, the package is installed from a "source" file, it means that the package is recompiled. For some computing reasons, when compiling by default, **R** uses the system (Windows) parameters (which doesn't have necessarily have the correct Java version).

There are two solutions:

- Compile the package by installing from the binary file:

    ```{r, echo = TRUE, eval = FALSE}
    install.packages("RJDemetra", type = "binary")
    ```

- Specify that we want to use the local parameters:

    ```{r, echo = TRUE, eval = FALSE}
    install.packages("RJDemetra", type = "source", INSTALL_opts = "--no-multiarch")
    ```

ℹ️ More informations: `r colFmt("https://github.com/jdemetra/rjdemetra/wiki/Installation-manual", html_color)`


## The command `library("RJDemetra")`{.r} returns an error message

The package **RJDemetra** requires Java version 8 or higher to work. If another package has been loaded before **RJDemetra** via the function `library()`{.r} and which doesn't requires an updated Java version, then an old Java version will be used during all the session (**R** is refractory to in-session version change). In case of using **RJDemetra** in a program, you have to specify at the very beginning of the program that **R** must use Java version 8, via the command:

```{r, echo = TRUE, eval = FALSE}
# Where Java is installed
Sys.setenv(JAVA_HOME = "C:/Users/Software/Java17/jdk17")
```

or load **RJDemetra** first

```{r, echo = TRUE, eval = FALSE}
# At the beginning of program
library("RJDemetra")
```

Else you have to restart a new **R** session.


## `Error array index = -1`

The message of the type `Error array index = -1` tells that an auxiliary variable is not found. It can be calendar regressor as other user defined variables (Easter effect, PSO = pure seasonal outlier...).

