---
title: "Installing JDemetra+"
author: "Tanguy BARTHELEMY"
# date: "2024-04-04"
output:
  github_document: default
  html_document:
    df_print: paged
    number_sections: TRUE
    highlight: "arrow"
    theme: "readable"
    css: "style.css"
  pdf_document: 
    includes:
        in_header: 
            - ../xelatex-emoji/xelatexemoji.sty
    extra_dependencies: ["url"]
    number_sections: TRUE
    highlight: "arrow"
    latex_engine: xelatex
always_allow_html: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r, echo = FALSE, eval = TRUE}
html_color <- "#000cff"
windows_path_color <- "#006817"

color_format <- function(x, color, url = TRUE) {
    color_name <- deparse(substitute(color))
    output_format <- knitr::opts_knit$get("rmarkdown.pandoc.to")

    x <- x |> gsub(pattern = "\\", replacement = "\\\\", fixed = TRUE)

    if (output_format == "latex") {
        x <- x |> gsub(pattern = "#", replacement = "\\#", fixed = TRUE)
        ret <- paste("\\textcolor{", color_name, ifelse(url, "}{\\url{", "}{\\nolinkurl{"), x, ifelse(url, "}}", "}}"), sep = "")
    } else if (output_format == "html") {
        ret <- paste("<font color='", color, "'>", x, "</font>", sep = "")
    } else if (output_format == "gfm+tex_math_dollars-yaml_metadata_block") {
        if (url) {
            ret <- paste("[", x, "](", x, ")", sep = "")
        } else {
            ret <- x |> gsub(pattern = "\\", replacement = "\\\\", fixed = TRUE)
            ret <- paste("${\\textsf{\\color{", color, "}{", ret, "}}}$", sep = "")
        }
    } else {
        ret <- x
    }

    return(ret)
}
```

```{r, echo = FALSE, eval = TRUE}
html_color <- "#000cff"
windows_path_color <- "#006817"
```

\definecolor{html_color}{HTML}{000cff}
\definecolor{windows_path_color}{HTML}{006817}

\renewcommand{\xelatexemojipath}[1]{../xelatex-emoji/images/#1}

## Installating the Graphical User Interface (GUI)

You can find the latest releases 

- in v2.x family: [here](https://github.com/jdemetra/jdemetra-app/releases)

In April 2024, the latest version is [v2.2.4](https://github.com/jdemetra/jdemetra-app/releases/tag/v2.2.4)

Scroll down on the page, download and unzip the file *jdemetra-2.2.4-bin.zip*

To launch the application execute the file **nbdemetra64.exe** located in the following sub folder: "... \jdemetra-2.2.4-bin\nbdemetra\bin"

- in v3.x family: [here](https://github.com/jdemetra/jdplus-main/releases)

In April 2024, the latest version is ([v3.2.2](https://github.com/jdemetra/jdplus-main/releases/tag/v3.2.2))

Scroll down on the page, download and unzip the file if you use windows *jdemetra-standalone-3.2.2-windows-x86_64.zip*

To launch the application execute the file **nbdemetra64.exe** located in the following sub folder: "... \jdemetra-standalone-3.2.2-windows-x86_64\nbdemetra\bin"

Version 3.x requires Java 17 or superior, *jdemetra-standalone-3.2.2-windows-x86_64.zip* contains a portable version of java 21, so you don't have to tackle this issue on your computer.

R packages related to version 3.x (rjd3...) also require Java 17 or superior, you can (and should) use the portable version provided with the Graphical user interface  to run them, this is explained [here](#Rconfig)

## Installing additionnal plug-ins

Extending...


## Installing the cruncher

To install the installer the cruncher :

- in v2, go on the page of [**JWSACruncher**](https://github.com/jdemetra/jwsacruncher) and **Download** and **unzip** the compressed folder `jwsacruncher-2.2.4-bin.zip`
- in v3, go on the page of [**jdplus-main**](https://github.com/jdemetra/jdplus-main) and **Download** and **unzip** the compressed folder `jwsacruncher-standalone-3.2.2-windows-x86_64.zip`


## Configaration to run rjd3 packages {#Rconfig}

Configuration of R and Rstudio

To use the packages in v3 in R, you need to specify the path to Java 17.

You need to execute this line at the beginning of your programs (with the correct path where you installed JDemetra+):

```{r, echo = TRUE, eval = FALSE}
Sys.setenv("JAVA_HOME" = "path/to/nbdemetra-3.2.2/jdk-21.0.2+13-jre")
```

