---
title: "Configuring JDemetra+ and related R tools"
author: "Tanguy BARTHELEMY"
# date: "2024-04-05"
output:
  github_document: default
  html_document:
    df_print: paged
    number_sections: TRUE
    highlight: "arrow"
    theme: "readable"
    css: "../style.css"
  pdf_document: 
    includes:
        in_header: 
            - ../../xelatex-emoji/xelatexemoji.sty
    extra_dependencies: ["url"]
    number_sections: TRUE
    highlight: "arrow"
    latex_engine: xelatex
always_allow_html: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r, echo = FALSE, eval = TRUE}
html_color <- "#000cff"
windows_path_color <- "#006817"

color_format <- function(x, color, url = TRUE) {
    color_name <- deparse(substitute(color))
    output_format <- knitr::opts_knit$get("rmarkdown.pandoc.to")

    x <- x |> gsub(pattern = "\\", replacement = "\\\\", fixed = TRUE)

    if (output_format == "latex") {
        x <- x |> gsub(pattern = "#", replacement = "\\#", fixed = TRUE)
        ret <- paste("\\textcolor{", color_name, ifelse(url, "}{\\url{", "}{\\nolinkurl{"), x, ifelse(url, "}}", "}}"), sep = "")
    } else if (output_format == "html") {
        ret <- paste("<font color='", color, "'>", x, "</font>", sep = "")
    } else if (output_format == "gfm+tex_math_dollars-yaml_metadata_block") {
        if (url) {
            ret <- paste("[", x, "](", x, ")", sep = "")
        } else {
            ret <- x |> gsub(pattern = "\\", replacement = "\\\\", fixed = TRUE)
            ret <- paste("${\\textsf{\\color{", color, "}{", ret, "}}}$", sep = "")
        }
    } else {
        ret <- x
    }

    return(ret)
}
```

```{r, echo = FALSE, eval = TRUE}
html_color <- "#000cff"
windows_path_color <- "#006817"
```

\definecolor{html_color}{HTML}{000cff}
\definecolor{windows_path_color}{HTML}{006817}

\renewcommand{\xelatexemojipath}[1]{../xelatex-emoji/images/#1}


## Context {-}

This document complements the [quick installation guide] (https://github.com/TanguyBarthelemy/JD-Tutorials/blob/master/Sheets/JDemetra+_quick_installation_guide_EN.pdf) by providing more details on the configuration required to run JDemetra+ and related tools.

On computers without administrator rights (*professional computer, for example*), it is recommended to create a folder **Software** under `r color_format("C:\\Users\\...\\Software", color = windows_path_color, url = FALSE)` or directly under `r color_format("C:\\Users\\Software", color = windows_path_color, url = FALSE)` where all software will be installed.

⚠️ Warning: when we specify an **absolute** path for a software (JDemetra+, Java, **R**, ...) in a program, a shortcut, a variable, ..., it must be modified each time any root repository is moved.

## Installation of JDemetra+ Graphical user interface

To install [JDemetra+ Graphical User Interface](https://github.com/jdemetra/jdemetra-app) see the [quick installation guide] (https://github.com/TanguyBarthelemy/JD-Tutorials/blob/master/Sheets/JDemetra+_quick_installation_guide_EN.pdf)

## Installation of the cruncher

To install [JDemetra+ cruncher](https://github.com/jdemetra/jdemetra-app) see the [quick installation guide] (https://github.com/TanguyBarthelemy/JD-Tutorials/blob/master/Sheets/JDemetra+_quick_installation_guide_EN.pdf)


## Installation of **R** and Rstudio

The JDemetra+ algorithms can be accessed in **R** via **R** packages.
To use **R**, it is better to use an IDE like Rstudio. All the executable files to download are under `r color_format("https://posit.co/download/rstudio-desktop/#download", color = html_color, url = TRUE)`.


### Installation of **R**

To install **R**, you should:

-   **Download** the binary file `R-4.3.2-win.exe` under `r color_format("https://cran.rstudio.com/bin/windows/base/", color = html_color, url = TRUE)`
-   **Execute** the executable to parameter and install **R**.


### Installation of Rstudio

**Download** the last Rstudio version (under `r color_format("https://posit.co/download/rstudio-desktop/#download", color = html_color, url = TRUE)`) and the **installer**.

If the installation via the file `.exe` fails (because it requires higher rights (administrator, elevation, ...), we will get a portable version of the Software. To do this:

- **Download** and **unzip** the compressed folder `.zip` in a folder named "Rstudio" (under `r color_format("C:\\Users\\Software", color = windows_path_color, url = FALSE)`) 
- **Create a shortcut** of the file `rstudio.exe` on the Desktop.


### Installation of **R** packages

To install a **R** package, there are several methods:

- Either it is on CRAN and you can install it directly with the function `install.packages()`{.r}
- Or it is on Github and you can install it directly with the function `install_github()`{.r} from the package **remotes**
- Or you have to retrieve the package from a folder (binary format) (`.zip`) and then install it with the function `install.packages()`{.r} with the arguments `repos = NULL, type = "binary"`.


#### R packages for JDemetra+ version 2.2.4

The packages in version 2 are:

| Name | Available on CRAN | Available on AUS | Github link |
|---|--|--|-----|
|     RJDemetra | ✔️ | ✔️ |      `r color_format("https://github.com/jdemetra/rjdemetra", color = html_color, url = TRUE)` |
|  rjdworkspace | ❌ | ✔️ | `r color_format("https://github.com/InseeFrLab/rjdworkspace", color = html_color, url = TRUE)` |
|    JDCruncheR | ❌ | ✔️ |      `r color_format("https://github.com/InseeFr/JDCruncheR", color = html_color, url = TRUE)` |
| rjwsacruncher | ✔️ | ✔️ |      `r color_format("https://github.com/AQLT/rjwsacruncher", color = html_color, url = TRUE)` |
|   rjdmarkdown | ✔️ | ✔️ |        `r color_format("https://github.com/AQLT/rjdmarkdown", color = html_color, url = TRUE)` |

The packages installation code is below:

```{r, echo = TRUE, eval = FALSE}
# If remotes is not installed
# install.packages("remotes")

install.packages("RJDemetra")
install.packages("rjwsacruncher")
install.packages("rjdmarkdown")

remotes::install_github("InseeFrLab/rjdworkspace")
remotes::install_github("InseeFr/JDCruncheR")

# Under AUS and on Insee computers
install.packages("rjdworkspace", repos = "https://nexus.insee.fr/repository/r-public")
install.packages("JDCruncheR", repos = "https://nexus.insee.fr/repository/r-public")
```


#### R packages for JDemetra+ version 3.x

Currently version 3 packages are NOT on CRAN. To install them, you need to go through Github:

```{r, echo = TRUE, eval = FALSE}
# If remotes is not installed
# install.packages("remotes")

remotes::install_github("rjdemetra/rjd3toolkit")

remotes::install_github("rjdemetra/rjd3x13")
remotes::install_github("rjdemetra/rjd3tramoseats")

remotes::install_github("rjdemetra/rjd3providers")
remotes::install_github("rjdemetra/rjdemetra3")

remotes::install_github("rjdemetra/rjd3filters")
remotes::install_github("rjdemetra/rjd3x11plus")

remotes::install_github("rjdemetra/rjd3sts")
remotes::install_github("rjdemetra/rjd3highfreq")
remotes::install_github("rjdemetra/rjd3stl")

remotes::install_github("rjdemetra/rjd3revisions")
remotes::install_github("rjdemetra/rjd3bench")
remotes::install_github("rjdemetra/rjd3nowcasting")

remotes::install_github("AQLT/ggdemetra3")
```


## Environment variables {#var_env}


### In Rstudio

To add environment variables in Rstudio, you need to add the variable's name and the variable's value.
There are 2 ways to fill it:

- Using the `.Renviron` file:

    - **Launch** the code `file.edit("~/.Renviron")`{.r} or `usethis::edit_r_environ()`{.r} (with  the package **usethis** and the argument `scope` which egals `"user"`{.r} or `"project"`{.r} if you are in a R project)
    - **Add** the variables to the file (with new lines)
    - **Save** the file

- Using the `.Rprofile` file:

    - **Launch** the code `file.edit("~/.Rprofile")`{.r} or `usethis::edit_r_profile()`{.r} (with  the package **usethis** and the argument `scope` which egals `"user"`{.r} or `"project"`{.r} if you are in a R project)
    - **Add** the variables with the fonction `Sys.setenv()`{.r} (with new lines)
    - **Save** the file



### PATH

The environment variable `PATH` in **R** is used to indicate to **R** where to find the executable files.

When you install a new software (for example JDemetra+, Rtools, Java...) that Rstudio uses, you have to modify this environment variable:

- **Get** the actual value of the variable `PATH` via the **R** command `Sys.getenv("PATH")`{.r} (Rstudio returns a succession of addresses as `C:/WINDOWS/system32;C:/WINDOWS`)

- **Copy paste** this value after `PATH = ` and add the paths towards the folder `r color_format("\\bin\\", color = windows_path_color, url = FALSE)` (binary) of the software newly installed, by separating them with semicolon (without space before or after).


    For the Rtools installation, the path is `r color_format("C:\\rtools42\\mingw64\\bin", color = windows_path_color, url = FALSE)` (depending on where Rtools was installed). You have to add `C:\\rtools42\\mingw64\\bin` or `C:/rtools42/mingw64/bin` (In **R**, `\` is a special character, so you have to replace the `\` by `/` or by `\\`). The path becomes `C:/WINDOWS/system32;C:/WINDOWS;C:/rtools42/mingw64/bin`.

- **Modify** the variable with the function `Sys.setenv()`{.r}. For Rtools, the command to launch is:

    ```{r, echo = TRUE, eval = FALSE}
    Sys.setenv(PATH = "C:/WINDOWS/system32;C:/WINDOWS;C:/rtools42/mingw64/bin")
    ```

ℹ️ NB: Generally a  32 bits version and a 64 bits version are available for downloading and installing a software. You need to check your processor type of your operating system to choose the right folder to download.

For this, you can launch the command:

```{r, echo = TRUE, eval = FALSE}
Sys.getenv("R_ARCH")
Sys.info()[["machine"]]
```

According to the result, the version is 32 bits or 64 bits : 

+------------+-------------+
| Version    | Output      |
+============+=============+
|            | /x64        |
| 64 bits    +-------------+
|            | x86-64      |
+------------+-------------+
+------------+-------------+
|            | /i386       |
| 32 bits    +-------------+
|            | x86_32      |
+------------+-------------+

More information on the variable `PATH` via the page `r color_format("https://java.com/fr/download/help/path.xml", color = html_color, url = TRUE)`.


## Check your set up

To ensure that everything works fine, you can launch some example of **R** code and check that there is no error:

```{r, echo = TRUE, eval = FALSE}
library("RJDemetra")

myseries <- ipi_c_eu[, "FR"]
x13_model <- x13(myseries) # X-13ARIMA method
ts_model <- tramoseats(myseries) # TRAMO-SEATS method

# Basic plot with the original series, the trend and the SA series
plot(x13_model, type_chart = "sa-trend")
```

To check the Java version we are using on **R**, you can try to install and use the package **rJava** and launch the command below:

```{r, echo = TRUE, eval = FALSE}
# If rJava is not installed
install.packages("rJava")
```

If the installation of **rJava** returns an error, it means that Java was incorrectly installed or incorrectly configured on **R**. You need to get back to the section [Environment variables](#var_env).

This block of code tests the Java version with which **R** works:

```{r, echo = TRUE, eval = FALSE}
library("rJava")
.jinit()
.jcall("java/lang/System", "S", "getProperty", "java.runtime.version")
```

Finally, you can display the Java version installed with which Windows works (it doesn't matter to us):

```{r, echo = TRUE, eval = FALSE}
system("java -version")
```


## Optional installations

Some supplementary installations are optional (that is they are no mandatory but bring external features):

-   [Miktek](https://miktex.org/howto/install-miktex) to produce PDF document with Latex
-   [Rtools](https://cran.r-project.org/bin/windows/Rtools/rtools42/rtools.html) to develop **R** packages and compile the code


## Problems you may encounter

### Problems installing **R** packages

If you get the following error while installing **R** packages:

```{r, echo = TRUE, eval = FALSE}
install.packages("RJDemetra")
```

```{r, error = TRUE, echo = FALSE, include = TRUE, eval = TRUE}
stop("Erreur: the chargement a échoué
Exécution arrêtée
*** arch - x64")
```

The problem doesn't come from Java but from the **R** package. By default, the package is installed from a "source" file, it means that the package is recompiled. For some computing reasons, when compiling by default, **R** uses the system (Windows) parameters (which doesn't have necessarily have the correct Java version).

There are two solutions:

- Compile the package by installing from the binary file:

    ```{r, echo = TRUE, eval = FALSE}
    install.packages("RJDemetra", type = "binary")
    ```

- Specify that we want to use the local parameters:

    ```{r, echo = TRUE, eval = FALSE}
    install.packages("RJDemetra", type = "source", INSTALL_opts = "--no-multiarch")
    ```

ℹ️ More information: `r color_format("https://github.com/jdemetra/rjdemetra/wiki/Installation-manual", color = html_color, url = TRUE)`


### The command `library("RJDemetra")`{.r} returns an error message

The package [**{RJDemetra]{.RPackages} requires Java version 8 or higher to work. If another package has been loaded before [**{RJDemetra}**]{.RPackages}  via the function `library()`{.r} and which doesn't requires an updated Java version, then an old Java version will be used during all the session (**R** is refractory to in-session version change). In case of using [**{RJDemetra}**]{.RPackages}  in a program, you have to specify at the very beginning of the program that **R** must use Java version 8, via the command:

```{r, echo = TRUE, eval = FALSE}
# Where Java is installed
Sys.setenv(JAVA_HOME = "C:/Users/Software/Java17/jdk17")
```

or load [**{RJDemetra}**]{.RPackages} first

```{r, echo = TRUE, eval = FALSE}
# At the beginning of program
library("RJDemetra")
```

Otherwise you have to restart a new **R** session.


### `Error array index = -1`

The message of the type `Error array index = -1` tells that an auxiliary variable is not found. It can be calendar regressor or other user defined variables (Easter effect, PSO = pure seasonal outlier...).


### The function `cruncher_and_param(...)`{.r} of the [**{JDCruncheR}**]{.RPackages} package returns an error

When you use the function `cruncher_and_param(...)`{.r} of the [**{JDCruncheR}**]{.RPackages} package, you can get the following error:

```{r, error = TRUE, echo = FALSE, include = TRUE, eval = TRUE}
stop("Error in cruncher(workspace = workspace, cruncher_bin_directory = cruncher_bin_directory,  : 
  There is an error in the path to the cruncher bin folder")
```

That means that the path to the cruncher is incorrectly configured. To solve this, you need to specify to R the path to the cruncher at the start of the program withe the function `options(...)`{.r} :

```{r, echo = TRUE, eval = FALSE}
options(cruncher_bin_directory = "C:/Users/Software/jwsacruncher-2.2.4-bin/bin")
```

To check that the path is valid, you can use the function `getOption(...)`{.r} :

```{r, echo = TRUE, eval = FALSE}
getOption("cruncher_bin_directory")
```
